---
title: "<b>Omix</b> - Integrative analysis of Multi-omics data in R"
author: "This vignette was created by Eléonore Schneegans and Nurun Fancy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{<b>Omix</b> - Integrative analysis of Multi-omics data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
  wrap: 72
---
```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
                        message = FALSE)
options(knitr.duplicate.label = "allow")
```

```{r setup}
library(Omix)
```

## Overview

The goal of Omix is to provide tools in R to build a complete analysis workflow 
for integrative analysis of data generated from multi-omics platform.

-   Quality control of single-omics data.

-   Formatting, normalisation, denoising of single-omics data.

-   Separate single-omics analysis.

-   Integration of multi-omics data for combined analysis. 

-   Publication quality plots and interactive analysis reports based of shinyApp.

## Running *Omix*

The `Omix` pipeline requires the following input:

-   rawdata_rna: a data-frame containing raw RNA count with rownames as gene and colnames as samples.

-   rawdata_protein:  a data-frame containing protein abundance with rownames as proteins and colnames as samples.

-   map_rna: A data-frame of two columns named `primary` and `colname` where primary should contain unique sample name with a link to sample metadata and colname is the column names of the `rawdata_rna` data-frame.

-   map_protein: A data-frame of two columns named `primary` and `colname` where primary should contain unique sample name with a link to sample metadata and colname is the column names of the `rawdata_protein` data-frame. 

-   metadata_rna: A data-frame containing `rna` assay specific metadata where rownames are same as the colnames of `rawdata_rna` data.frame.

-   metadata_protein: A data-frame containing `protein` assay specific metadata where rownames are same as the colnames of `rawdata_protein` data.frame.

-   sample_metadata: A data-frame containing sample level metadata for both omics assays.


### Step one - Generate MultiAssayExperiment object

Call required packages for vignette:
```{r}
# library(Omix)
```

