---
title: "<b>Omix</b> - Integrative analysis of Multi-omics data in R"
author: "This vignette was created by Eléonore Schneegans and Nurun Fancy"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{<b>Omix</b> - Integrative analysis of Multi-omics data in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
  wrap: 72
---
```{r style, echo=FALSE, results='asis', message=FALSE}
BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
                        message = FALSE)
options(knitr.duplicate.label = "allow")
```

```{r setup}
library(Omix)
```

## Overview

The goal of Omix is to provide tools in R to build a complete analysis workflow 
for integrative analysis of data generated from multi-omics platform.

-   Generate a multi-omics object using `MultiAssayExperiment`.

-   Quality control of single-omics data.

-   Formatting, normalisation, denoising of single-omics data.

-   Separate single-omics analyses.

-   Integration of multi-omics data for combined analysis. 

-   Publication quality plots and interactive analysis reports based of shinyApp.

## Running *Omix*

The `Omix` pipeline requires the following input:

-   rawdata_rna: a data-frame containing raw RNA count with rownames as gene and colnames as samples.

-   rawdata_protein:  a data-frame containing protein abundance with rownames as proteins and colnames as samples.

-   map_rna: A data-frame of two columns named `primary` and `colname` where primary should contain unique sample name with a link to sample metadata and colname is the column names of the `rawdata_rna` data-frame.

-   map_protein: A data-frame of two columns named `primary` and `colname` where primary should contain unique sample name with a link to sample metadata and colname is the column names of the `rawdata_protein` data-frame. 

-   metadata_rna: A data-frame containing `rna` assay specific metadata where rownames are same as the colnames of `rawdata_rna` data.frame.

-   metadata_protein: A data-frame containing `protein` assay specific metadata where rownames are same as the colnames of `rawdata_protein` data.frame.

-   sample_metadata: A data-frame containing sample level metadata for both omics assays.


### Step one - Generate MultiAssayExperiment object

Call required packages for vignette:
```{r}
library(Omix)
```

First we must get the data from Omix for the vignette:

```{r}
rawdata_rna <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/rawdata_rna.rds')
rawdata_protein <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/rawdata_protein.rds')
map_rna <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/map_rna.rds')
map_protein <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/map_protein.rds')
metadata_rna<- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/metadata_rna.rds')
metadata_protein <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/metadata_protein.rds')
sample_metadata <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/sample_metadata.rds')
rna_qc_data_matrix <- readRDS('~/RDS_ukdri/multiomics/Omix_test_env/data/rna_qc_data_matrix.rds')
```

Input files structure

# Raw rna counts

Raw RNA is a data frame of raw counts, with features as rows and samples as columns

```{r}
print(rawdata_rna)
```

# Raw protein abundance

Raw protein is a data frame of raw counts, with features as rows and samples as columns

```{r}
print(rawdata_protein)
```
# Maps

Maps are data frame containing two columns: `primary` and `colname`. 
The `primary` column should be the individual ID the individual metadata, and the 
`colname` the matched sample names from the raw matrices columns. If sample and
individual ids are the same, maps aren't needed (primary and colnames are the same).
```{r}
print(map_protein)
print(map_rna)
```
# Technical metadata

Technical metadata are data frames contain the column `colname` that should match
the sample names in the raw matrices, and any additional columns related to technical artefacts like batches 
```{r}
print(metadata_rna)
```
# Sample metadata

Sample metadata should contain individual level metadata, where one column matches the `primary` column in maps.
```{r}
print(sample_metadata)
```
# Optional inputs

Optional inputs contain additional data frame used for QC visualisation purposes 
```{r}
print(rna_qc_data_matrix)
```


# Generate a multiassay experiment object 

To run Omix, we first need to generate a multi-omics object

```{r}
multiassay=generate_multiassay(rawdata_rna = rawdata_rna,
                  rawdata_protein = rawdata_protein,
                  individual_to_sample=FALSE,
                  map_rna = map_rna,
                  map_protein = map_protein,
                  metadata_rna = metadata_rna,
                  metadata_protein = metadata_protein,
                  individual_metadata = individual_metadata,
                  map_by_column = 'individualID',
                  rna_qc_data=TRUE,
                  rna_qc_data_matrix=rna_qc_data_matrix,
                  organism='human')
          
```

The MultiAssayExperiment object was succesfully created. 
The following steps will process and perform QC on each omic layers of the 
`multiassay` object.

### Step two - Process transcriptomics data
```{r}
multiassay=process_rna(multiassay=multiassay,
            transformation='vst',
            protein_coding=TRUE,
            min_count = 10,
            min_sample = 0.6,
            dependent= 'diagnosis',
            levels=c('Control','Mid','Late'),
            covariates=c('pmi','msex','age_death'),
            filter=TRUE,
            batch_correction=TRUE,
            batch="libraryBatch",
            remove_sample_outliers= TRUE)
```



### Optional - Add custom processed transcriptomics slot
```{r}
multiassay=processed_rna(multiassay=multiassay,
                         custom_processed_df=custom_rna)
```



### Step three - Process proteomics data


```{r}
multiassay=process_protein(
    multiassay=multiassay,
    min_sample = 0.5,
    dependent =  "diagnosis",
    levels = c('Control','Mid','Late'),
    imputation = 'distribution',
    remove_feature_outliers= FALSE,
    batch_correction= TRUE,
    batch="Batch",
    correction_method="Median_centering",
    remove_sample_outliers=TRUE,
    denoise=TRUE,
    covariates=c('pmi','msex','age_death'))
```


### Optional - Add custom processed transcriptomics slot
```{r}
multiassay=processed_protein(multiassay=multiassay,
                         custom_processed_df=custom_protein)
```

# Processed multiassay object

```{r}
print(multiassay)
```

```{r}
upsetSamples(multiassay)
```

### Step Four - Generate QC dashboards 


### Step Five - Single omic analyses 

```{r}
multiassay=protein_single_analysis(multiassay,
                         slot="protein_processed",
                         dependent='diagnosis',
                         covariates=c('pmi','msex','age_death'),
                         levels=c('Control','Mid','Late'))
```


```{r}
multiassay@metadata[["DEP"]][["plot"]][["LatevsControl"]]
```


```{r}
multiassay=rna_single_analysis(multiassay,
                    dependent='diagnosis',
                    levels=c('Control','Mid','Late'))
```

```{r}
multiassay@metadata[["DEG"]][["plot"]][["LatevsControl"]]
```

### Step Five - Multi-omics integration using DIABLO

```{r}
multiassay=vertical_integration(multiassay,
                                 slots=c('rna_processed',
                                         'protein_processed'),
                                 dependent='Group',
                                 map_by_column='individualID',
                                 integration='DIABLO',
                                 ncomp=2,
                                 range=seq(5,100,by=10))
```

### Step Five - Multi-omics integration using MOFA

```{r}
multiassay=vertical_integration(multiassay,
                                 slots=c('rna_processed',
                                         'protein_processed'),
                                 integration='MOFA')
```


### Step Six - Multi-omics analyses 
```{r}
int_results=integrative_results-supervised(multiassay,
                                integration = "DIABLO",
                                component = 1,
                                correlation_threshold=0.5,
                                disease_id='MONDO_0004975')
```

