---
title: "<b>Omix</b> - Single Omic analyses report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: "flatly"
    toc: true
    toc_float: true
    highlight: tango
    fig_caption: true
    keep_md: false
    css: !expr system.file("rmarkdown/css/style.css", package = "Omix")
    includes:
      in_header: header.html
      after_body: !expr system.file("rmarkdown/html/footer.html", package = "Omix")
params:
  metadata_path: false
---

```{r style, echo=FALSE, results='asis', message=TRUE, warning=FALSE}
library(DT)
library(utils)
library(knitr)
library(htmltools)
library(plotly)
library(dplyr)
library(Omix)
library(table1)


BiocStyle::markdown()
knitr::opts_chunk$set(tidy = FALSE,
                      message = FALSE,
                      warning = FALSE)
options(knitr.duplicate.label = "allow")
```


```{r, load_data, results='hide', include=FALSE}
data <- qs::qread(params$metadata_path)
list2env(data, envir = environment())
slot=param$slot
database=param$database
num_path=as.numeric(param$num_path)

dependent=parameters_processing_protein$dependent
design_prot=paste(as.character(parameters_analysis_protein$design), collapse = '')
```

## Single omic report overview 

* **Cohort overview**
* **Proteomics**
* **Functional enrichment**


### Analysis summary

**Analysis name:** `r slot`  
**Enrichment database:** `r database`  

Cohort overview
=======================================================================

## Table1

```{r, echo=FALSE}
dependent_var=metadata[,dependent]
table1::table1(~. | dependent_var, data=metadata[,c(clinical_covariates)])

```


Proteomics
=======================================================================


```{r echo=FALSE}
if(parameters_processing_protein$denoise==TRUE){
exp=paste(parameters_processing_protein$covariates,collapse = ' + ')
ans=paste(design_prot,'+',exp)
}else{ ans= 'No denoising, design is shown above'}
```

## Analysis parameters 

**Dependent variable:** `r parameters_analysis_protein$dependent`   
**Differential expression design:** `r design_prot`   
**Actual design (if denoising in processing step):** `r ans`   
**Log2 Fold Change threshold:** `r parameters_analysis_protein$log2FoldChange`  
**Adjusted p value threshold:** `r parameters_analysis_protein$padj`    


## Differential expression {.tabset}


### Results

```{r, echo=FALSE}

  DT::datatable(DEP[[slot]],
                rownames = FALSE,
                escape = FALSE,
                options = list(scrollX=T,
                               autoWidth = TRUE,
                               dom = 'Blfrtip'))

```

### Volcano plot

```{r, echo=FALSE}

  plot=volcano_interactive(DEP[[slot]], padj=0.05)
  plot

```


Functional enrichment results  
=======================================================================

## Proteomics {.tabset}

### Up regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$up)){
  cat( 'No pathways were found to be significantly up regulated')
}else{
  DT::datatable(DEP$functional_enrichment[[slot]]$up[[database]],
                rownames = FALSE,
                escape = FALSE,
                options = list(scrollX=T,
                               autoWidth = TRUE,
                               dom = 'Blfrtip'))
} 

```

### Down regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$down)){
  cat( 'No pathways were found to be significantly down regulated')
}else{
  DT::datatable(DEP$functional_enrichment[[slot]]$down[[database]],
                rownames = FALSE,
                escape = FALSE,
                options = list(scrollX=T,
                               autoWidth = TRUE,
                               dom = 'Blfrtip'))
} 

```

Functional enrichment plot
=======================================================================

## Proteomics {.tabset}

### Up regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$up)){
  cat( 'No pathways were found to be significantly up regulated')
}else{
  plot=DEP$functional_enrichment[[slot]]$up$plot[[database]]
  plot
} 

```

### Down regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$down)){
  cat( 'No pathways were found to be significantly down regulated')
}else{
  plot=DEP$functional_enrichment[[slot]]$down$plot[[database]]
  plot
} 

```

Functional enrichment results as networks
=======================================================================

## Networks 

### Up regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$up)){
  cat( 'No pathways were found to be significantly up regulated')
}else{
enr_path=DEP$functional_enrichment[[slot]]$up[[database]]
enr_path$genes=DEP$functional_enrichment[[slot]]$up[[database]]$Genes


visualise_pathway(pathways=enr_path,num_path=num_path)
} 

```

### Down regulated proteins

```{r, echo=FALSE}

if (is.null(DEP$functional_enrichment[[slot]]$down)){
  cat( 'No pathways were found to be significantly down regulated')
}else{
enr_path=DEP$functional_enrichment[[slot]]$down[[database]]
enr_path$genes=DEP$functional_enrichment[[slot]]$down[[database]]$Genes
visualise_pathway(pathways=enr_path,num_path=num_path)
} 

```

<br>Omix v`r utils::packageVersion("Omix")` -- `r Sys.time()`

